---
title: "Identifying non-neutral TCRb clonotypes in longitudinal PBL samples"
author: "Endika Prieto-Fernández"
date: "2025-04-01"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

**Description:**

A pipeline for filtering out non-neutral TCRb clonotype trajectories in the blood of ICB-treated patients is based on removing the natural variability of the TCR repertoire characterized in healthy individuals, and includes additional filters to remove background noise caused by technological bias. Once non-neutral clonotypes have been identified, distance matrix-based k means clustering groups the trajectories into distinct waves of expansion or contraction.

# lib

```{r, message=FALSE, warning=FALSE}

library(dplyr)
library(edgeR)
library(factoextra)
library(ggplot2)
library(ggrepel)
library(grid)
library(httr)
library(patchwork)
library(purrr)
library(RColorBrewer)
library(readxl)
library(scales)
library(stringr)
library(tidyverse)
library(zoo)

```

# fun

local

```{r}
source("/Users/endikaprieto/Library/CloudStorage/OneDrive-VHIO/pbl/bulk/tools/ibilbide/v6/auxiliary_functions_v6.R")
```

# var

load hd

```{r, warning=F}

#hd <- c("HD106_SEQTR", "HD197_SEQTR")                                                  # Gros et al. SEQTR
#hd <- c("HD106", "HD159", "HD196", "HD197", "HD199")                                   # Gros et al. adaptive.
#hd <- c("HD106_hsd", "HD159_hsd", "HD196_hsd", "HD197_hsd", "HD199_hsd")               # Gros et al. hsd.
#hd <- c("CHU01", "CHU02", "CHU03")                                                     # Chu et al. adaptive.
#hd <- c(paste0("MIT0", c(1:9)), paste0("MIT", c(10:25)))                               # Mitchell et al. adaptive.
#hd <- c("YOSD1_CD4", "YOSD2_CD4", "YOSD3_CD4", "YOSD4_CD4", "YOSD5_CD4", "YOSD6_CD4")  # Yoshida et al. cd4. adaptive.
#hd <- c("YOSD1_CD8", "YOSD2_CD8", "YOSD3_CD8", "YOSD4_CD8", "YOSD5_CD8", "YOSD6_CD8")  # Yoshida et al. cd8. adaptive.
#hd <- c("HD106_SEQTR", "HD197_SEQTR"
#        , "HDBR01_SEQTR", "HDBR02_SEQTR", "HDBR03_SEQTR", "HDBR06_SEQTR"
#        , "HD106bis_SEQTR_1e6", "HD197bis_SEQTR_1e6")                                   # SEQTR cohort, RNA-based.
hd <- c("HD106_SEQTR",
        "HD197_SEQTR",
        "HDBR01_SEQTR_ReSeq",
        "HDBR02_SEQTR_ReSeq",
        "HDBR03_SEQTR_ReSeq",
        "HDBR06_SEQTR_ReSeq")

load_data(samples = hd,
          datapath.pbl = "/Volumes/datos_lab/tcrb/data_blood/hd/",
          output.list.name = "healthy",
          technology = "SEQTR",
          min_freq = 1e-4
          )

#plot_mean_counts(healthy, count_type = "count")
p <- initial_data_qc(healthy, ncol = 4)

pdf("/Users/endikaprieto/Desktop/healthy_initial_qc.pdf", width = 25, height = 15)
p$plot
dev.off()

```
# pt

pt <- c("MSI001", "MoTriColor01", "MoTriColor03", "RIO008", "RIO009", "RIO025", "RIO029", "RIO069", "RIO136", "RIO142", "RIO208")
pt <- c("NOUS_05", "NOUS_09", "NOUS_12", "NOUS_15", "NOUS_16", "NOUS_18")
pt <- c("16097-101")
pt <- c("NOUS_18_SEQTR")

```{r, warning=FALSE}

pt <- c("NOUS_18_SEQTR")

load_data(samples = pt,
          datapath.pbl = "/Volumes/datos_lab/tcrb/data_blood/patients/",
          output.list.name = "patients",
          technology = "SEQTR",
          min_freq = min_freq
          )

plot_mean_counts(patients, count_type = "count")
initial_data_qc(patients, ncol = 1)

```

# bcam collab

Clonotype assignment and repertoire anonymization.

```{r}
# --- Function ---
add_clono_codes <- function(x) {
  if (!is.list(x) || length(x) == 0) {
    stop("'x' must be a non-empty list.")
  }
  x_out <- x
  mapping_list <- list()
  for (sample_name in names(x)) {
    sample_obj <- x[[sample_name]]
    if (!is.list(sample_obj) || length(sample_obj) == 0) {
      warning(sprintf("Skipping '%s': not a valid non-empty list.", sample_name))
      next
    }
    aa_all <- unlist(
      lapply(sample_obj, function(df) {
        if (!"aminoAcid" %in% colnames(df)) {
          stop(sprintf("Missing 'aminoAcid' in '%s'.", sample_name))
        }
        as.character(df$aminoAcid)
      }),
      use.names = FALSE
    )
    aa_unique <- unique(aa_all[!is.na(aa_all)])
    sample_map <- data.frame(
      aminoAcid = aa_unique,
      clono = paste0("clono_", seq_along(aa_unique)),
      stringsAsFactors = FALSE
    )
    mapping_list[[sample_name]] <- sample_map
    for (tp_name in names(sample_obj)) {
      df <- sample_obj[[tp_name]]
      idx <- match(as.character(df$aminoAcid), sample_map$aminoAcid)
      df$clono <- sample_map$clono[idx]
      cols_to_remove <- intersect(c("nucleotide", "aminoAcid"), colnames(df))
      df <- df[, setdiff(colnames(df), cols_to_remove), drop = FALSE]
      x_out[[sample_name]][[tp_name]] <- df
    }
  }
  return(list(
    data = x_out,
    mapping = mapping_list
  ))
}
# --- Save data for sharing ---
bcam.path <- "/Users/endikaprieto/Library/CloudStorage/OneDrive-VHIO/collabs/BCAM-BIlbao/VHIO-BCAM-Shared"
# --- Healthy donors ---
res_healthy <- add_clono_codes(healthy)

healthy_with_clono <- res_healthy$data
saveRDS(healthy_with_clono, file.path(bcam.path, "Healthy-Data-Anonymized.rds"))

healthy_clono_mapping <- res_healthy$mapping
saveRDS(healthy_clono_mapping, file.path(bcam.path, "Healthy-Clono-Key.rds"))
# --- Patients ---
res_patients <- add_clono_codes(patients)

patients_with_clono <- res_patients$data
saveRDS(patients_with_clono, file.path(bcam.path, "Patients-Data-Anonymized.rds"))

patients_clono_mapping <- res_patients$mapping
saveRDS(patients_clono_mapping, file.path(bcam.path, "Patients-Clono-Key.rds"))
```

# norm

```{r}

norm_data <- normalize_data_edgeR(healthy_data = "healthy",
                                  patients_data = "patients")

healthy <- norm_data[hd]
patients <- norm_data[pt]

plot_mean_counts(norm_data, count_type = "count")
plot_mean_counts(norm_data, count_type = "count.norm")

```

calculate variability

```{r, warning=FALSE, fig.width=12, fig.height=6}

calculate_variability(input.data = "healthy")

```

# stats

calculate p-values

```{r, warning=F}

day.limit = NULL

calculate_statistics(input.data = "patients",
                     day.limit = day.limit,
                     p.val = 0.05
                     )
```

# traj

identify non-neutral trajectories

```{r, warning=FALSE, fig.width=12, fig.height=6}

identify_trajectories(input.data = "patients",
                      day.limit = day.limit, 
                      true.window = TRUE,
                      filter.dropouts = FALSE,
                      filter.na.valley = FALSE, na.valley = 4,
                      traj.filter = FALSE
                      )
```

# k

run following chunks for each pt

```{r, warning=F}

sample = "NOUS_18_SEQTR"

calculate_k_means(input.data = "patients",
                  sample = sample,
                  day.limit = day.limit,
                  min_freq = min_freq
                  )
```

# clust

```{r, warning=FALSE, fig.width= 20}

separate_trajectories(input.data = "patients",
                      sample = sample,
                      max.freq = 2,
                      n.col = 5,
                      day.limit = day.limit,
                      k = 20
                      )
```

# high

```{r, warning=FALSE, fig.width= 20}

highligth_trajectories(input.data = "patients",
                       sample = sample,
                       day.limit = day.limit,
                       filter.traj = TRUE,
                       max.freq = 5,
                       top.n.clonotypes = 500,
                       plot.time = TRUE,
                       #icb.1 = c(0,12,42,56,84,112,140,154,166,182,196,210),
                       biopsy.day = 0,
                       datapath.annot = "/Volumes/datos_lab/tcrb/data_tumor", 
                       technology = "hsd",
                       )
```

# traj tum

```{r, warning=FALSE}

names(pbl_tumor[[sample]])
biopsy <- "p136-TILs_TCRB in frame"
max.biopsy = 1000

calculate_k_means(input.data = "pbl_tumor",
                  sample = sample,
                  weight.frequency = 0.2,
                  weight.jumps = 0.2,
                  weight.na.pattern = 0.1,
                  weight.contractions = 0.5,
                  day.limit = day.limit,
                  biopsy = biopsy,
                  mode = "fold-change",
                  max.biopsy = max.biopsy
                  )
```


```{r, warning=FALSE, fig.width= 20}

separate_trajectories(input.data = "pbl_tumor",
                      sample = sample,
                      max.freq = 10,
                      day.limit = NULL,
                      biopsy = biopsy,
                      k = 10
                      )

```


```{r, warning=FALSE, fig.width=16, fig.height=8}

highligth_trajectories(input.data = "pbl_tumor",
                       sample = sample,
                       biopsy = biopsy,
                       filter.traj = TRUE,
                       max.biopsy = max.biopsy,
                       day.limit = day.limit
                       )
```

#capture

Internal control: number of clonotypes found in biopsies between trajectory or randomly selected clonotypes in blood. Is Bilobide enriched in PBL-TIL?

```{r, message=FALSE, warning=FALSE}

count_clonotypes(sample = sample, n_iter = 25)

```

# save

```{r}

save_data(datapath = "/Users/endikaprieto/Desktop",
          export.raw.data = TRUE,
          dpi = 200)

```

# add-on

Trajectories of the TCR library clonotypes.

```{r, message=FALSE, warning=FALSE}

# non-neutral clonotypes:
path <- "/Users/endikaprieto/Library/CloudStorage/OneDrive-VHIO/pbl/bulk/tools/tcr_beta_dynamics_analysis-main_EP/analyses/360RIO/results/RIO136/pbl_non_neutral_filtered_RIO136.tsv"
data <- read.table(path, sep = "\t", header = TRUE) %>%
  mutate(TRBV = str_extract(nucleotide, "TCRBV[0-9]+-0?[0-9]"),
         TRBV = gsub("-0", "-", TRBV),
         TRBV = gsub("TCRBV0", "TCRBV", TRBV),
         TRBV = gsub("TCRBV", "TRBV", TRBV),
         TRBJ = str_extract(nucleotide, "TCRBJ[0-9]+-0?[0-9]"),
         TRBJ = gsub("-0", "-", TRBJ),
         TRBJ = gsub("TCRBJ0", "TCRBJ", TRBJ),
         TRBJ = gsub("TCRBJ", "TRBJ", TRBJ),
  ) %>%
  mutate(trb = paste0(aminoAcid, "_", TRBV, "_", TRBJ))

# viral info, RIO136 HLA-restricted:
path <- "/Users/endikaprieto/Library/CloudStorage/OneDrive-VHIO/tcr_pbl/single_cell/sc_pipeline/patients/RIO136/outs/vdjdb/vdjdb_rio136_trb.tsv"
vdjdb <- read_tsv(path, col_names = TRUE) %>%
  as.data.frame() %>%
  select(2:13) %>%
  mutate(
    V = sub("\\*.*", "", V),
    J = sub("\\*.*", "", J)
  ) %>%
  mutate(TCRb = paste0(CDR3, "_", V, "_", J)) %>%
  select(-CDR3, -V, -J, -Gene, -Species)
data <- data %>% left_join(vdjdb, by = "TCRb")

# Tested TCR:
path <- "/Volumes/labagros/pierre/proj/rootpath/TCR_library_screen/RIO136/TCRlib_rio136/TXTO_TCRTable_VHIOwv1_5Criteria_1001TCRs_20201110_154539.tsv"
tcr_lib <- read_tsv(path, col_names = FALSE) %>%
  mutate(X14 = paste0("TRA:", X3, ";TRB:", X6)) %>%
  select(X1, X2, X3, X4, X5, X6, X7, X14) %>%
  as.data.frame()
colnames(tcr_lib) <- c("id", "trav", "tra", "traj", "trbv", "trb", "trbj", "tcr")
tcr_lib <- tcr_lib %>%
  mutate(
    trav = sub("\\*.*", "", trav),
    traj = sub("\\*.*", "", traj),
    trbv = sub("\\*.*", "", trbv),
    trbj = sub("\\*.*", "", trbj),
  ) %>%
  mutate(
    TCRb = paste0(trb, "_", trbv, "_", trbj),
    TCRa = paste0(tra, "_", trav, "_", traj)
  )

## hits
hits <- read.table("/Users/endikaprieto/Library/CloudStorage/OneDrive-VHIO/tcr_pbl/tcr_libraries/analysis/rio136_tcr_pipeline_ratio/results/TCRlib_all_hits.txt")$V1
hits <- gsub("VHIOwv1_", "", hits)
tcr_lib$hits <- ifelse(tcr_lib$id %in% hits, "Y", "N")

## validation
validated <- paste0("ChosenID.", c(12, 143, 146, 159, 16, 187, 195, 214, 23, 24, 232, 295, 31, 440, 445, 525, 531, 572, 618, 60, 640, 83, 94))
unknown <- paste0("ChosenID.", c(347, 484, 684))
tcr_lib$reactive <- ifelse(tcr_lib$id %in% validated, "Y", "unknown")
tcr_lib$reactive <- ifelse(tcr_lib$id %in% unknown, "N", tcr_lib$reactive)

## private or shared
private <- paste0("ChosenID.", c(12, 146, 159, 214, 23, 684, 484, 525, 531, 618, 347, 24))
shared <- paste0("ChosenID.", c(143, 16, 195, 295, 31, 232, 187, 640, 440, 572, 60, 83, 94, 445))
tcr_lib$type <- ifelse(tcr_lib$id %in% private, "private", "unknown")
tcr_lib$type <- ifelse(tcr_lib$id %in% shared, "shared", tcr_lib$type)
tcr_lib <- tcr_lib %>%
  select(TCRb, id, hits, reactive, type) %>%
  group_by(TCRb, hits, reactive, type) %>%
  summarize(id = paste(id, collapse = ","), .groups = "drop") %>%
  distinct()
data <- data %>% left_join(tcr_lib, by = "TCRb")

# complete data
metadata <- data %>% select(-time, -freq_in, -log10value, -count) %>% distinct()
timepoints <- unique(data$time)
data <- data %>%
  select(nucleotide, time, freq_in, count) %>%
  complete(nucleotide,
           time = timepoints,
           fill = list(freq_in = 0.0001, count = 0)
           ) %>%
  left_join(metadata, by = "nucleotide") %>% as.data.frame() %>% distinct()
data <- data %>% mutate(log10value = log10(freq_in))

# plot

long_data <- data %>% filter(traj %in% c(1, 5, 6, 7, 12, 27:30))
long_data <- data
timepoints <- length(unique(long_data$time))

count_traj <- long_data %>%
  group_by(traj) %>%
  summarise(
    count_total = n() / timepoints,
    .groups = "drop"
    )


plot <- ggplot() +
  
  # Ribbon for uncertainty (95% quantile range)
  geom_ribbon(data = long_data %>% 
                filter(nucleotide != "cumulative_frequency") %>%
                group_by(traj, time) %>%
                summarise(ymin = quantile(log10value, 0.05, na.rm = TRUE),
                          ymax = quantile(log10value, 0.95, na.rm = TRUE),
                          .groups = "drop"),
              aes(x = time, ymin = ymin, ymax = ymax),
              fill = "gray70", alpha = 0.25) +
  
  # Mean trajectory line
  geom_line(data = long_data %>% 
              filter(nucleotide != "cumulative_frequency") %>%
              group_by(traj, time) %>%
              summarise(mean_value = mean(log10value, na.rm = TRUE),
                        .groups = "drop"),
            aes(x = time, y = mean_value),
            color = "black", linewidth = 0.5, linetype = "dashed") +
  
  # Filtered lines where hits == "Y", painted by a variable (e.g., 'nucleotide')
  geom_line(data = long_data %>% 
              filter(long_data$p136.Tfrozen_TCRB.in.frame == "Y"),
            aes(x = time, y = log10value, color = nucleotide),
            size = 0.5, alpha = 0.7) +
  
  theme_minimal() +
  
  scale_x_continuous() +
  scale_y_continuous(
    breaks = log10(c(0.0001, 0.001, 0.01, 0.1, 1, 10, 100)),
    labels = c("0.0001", "0.001", "0.01", "0.1", "1", "10", "100"),
    limits = c(log10(0.0001), log10(100))
    ) +
  
  labs(y = "frequency") +
  facet_wrap(~traj, ncol = 5) +
  
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.2),
    aspect.ratio = 0.5, legend.position = "none"
    ) +
  
  # Add count labels for each trajectory
  geom_text(data = count_traj,
            aes(x = 25, y = 0.5, label = count_total),
            inherit.aes = FALSE,
            position = position_nudge(y = 0), size = 4)



```

Tag non-neutral clonotypes into CD4+ or CD8+ T cells.

```{r}

# Load single-cell data.

library(Seurat)

sc.data.cd8 <- readRDS("/Users/endikaprieto/Library/CloudStorage/OneDrive-VHIO/pbl/single_cell/sc_pipeline/patients/RIO136/seurat/int_cd8_annot.rds")@meta.data
cd8 <- unique(sc.data.cd8$trb)
sc.data.cd4 <- readRDS("/Users/endikaprieto/Library/CloudStorage/OneDrive-VHIO/pbl/single_cell/sc_pipeline/patients/RIO136/seurat/int_cd4_annot.rds")@meta.data
cd4 <- unique(sc.data.cd4$trb)

# Load non-neutral clonotyopes.

path <- "/Users/endikaprieto/Library/CloudStorage/OneDrive-VHIO/pbl/bulk/tools/tcr_beta_dynamics_analysis-main_EP/analyses/360RIO/results/RIO136/pbl_raw_RIO136.tsv"
non.neutral <- read.table(path, sep = "\t", header = TRUE) %>%
  mutate(TRBV = str_extract(nucleotide, "TCRBV[0-9]+-0?[0-9]"),
         TRBV = gsub("-0", "-", TRBV),
         TRBV = gsub("TCRBV0", "TCRBV", TRBV),
         TRBV = gsub("TCRBV", "TRBV", TRBV),
         TRBJ = str_extract(nucleotide, "TCRBJ[0-9]+-0?[0-9]"),
         TRBJ = gsub("-0", "-", TRBJ),
         TRBJ = gsub("TCRBJ0", "TCRBJ", TRBJ),
         TRBJ = gsub("TCRBJ", "TRBJ", TRBJ),
  ) %>%
  mutate(trb = paste0(aminoAcid, "_", TRBV, "_", TRBJ)) %>% select(-TRBV, -TRBJ)

compartment <- non.neutral %>%
  mutate(compartment = case_when(
    trb %in% cd4 ~ "cd4",
    trb %in% cd8 ~ "cd8",
    TRUE ~ "unknown"
  )) %>%
  select(nucleotide, compartment) %>%
  distinct()

non.neutral <- non.neutral %>% left_join(compartment, by = "nucleotide") %>% select(-trb)

cumulative <- non.neutral %>% select(nucleotide, time, freq_in, compartment)

cumulative.cd4 <- cumulative %>%
  filter(compartment == "cd4") %>%
  group_by(time) %>%
  summarise(total_freq = sum(freq_in)) %>%
  arrange(time)
write.table(cumulative.cd4, file.path(dirname(path), "cumulative_frequency_cd4.tsv"), sep = "\t", col.names = TRUE, row.names = FALSE)

cumulative.cd8 <- cumulative %>%
  filter(compartment == "cd8") %>%
  group_by(time) %>%
  summarise(total_freq = sum(freq_in)) %>%
  arrange(time)
write.table(cumulative.cd8, file.path(dirname(path), "cumulative_frequency_cd8.tsv"), sep = "\t", col.names = TRUE, row.names = FALSE)

cumulative.unknown <- cumulative %>%
  filter(compartment == "unknown") %>%
  group_by(time) %>%
  summarise(total_freq = sum(freq_in)) %>%
  arrange(time)
write.table(cumulative.unknown, file.path(dirname(path), "cumulative_frequency_unknown.tsv"), sep = "\t", col.names = TRUE, row.names = FALSE)

```

